Variation in site entropy explains differences in structure-sequence relationships of proteins

### 2015-06-09 manuscript/manuscript.tex
Adding citations to the manuscript

\usepackage[margin = 1.0 in]{geometry}
\usepackage{natbib}

@@

\title{Variation in site entropy explains differences in structure-sequence relationships of proteins}
\author{Eleisha L. Jackson$^{2^*}$, Amir Shahmoradi$^{1*}$, Claus O. Wilke$^{2^{**}}$}
\begin{document}

@@

\noindent
$^1$ Department of Physics, The University of Texas at Austin, Austin, TX 78712, USA \\
$^2$ Institute of Cellular and Molecular Biology, Center for Computational Biology and Bioinformatics, and Department of Integrative Biology, The University of Texas at Austin, Austin, Texas, 78712 USA\\

@@

\begin{abstract}
Recent work has shown that structural properties are capable of predicting site-specific sequence variability for a given protein. However, the strength and significance of these structure-sequence relations appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. There have two works that have investigated structural predictors of site variability which both present different results based on the strength of correlations of structural predictors. According to \cite{Yehetal2014}, both solvent accessibility and  local packing density are predictors of site-wise variability with local packing density being strongest predictor of site-wise variability. However, recently, \cite{Shahmoradietal2014} compared local packing density, residue flexibility, and solvent accessibility in viral proteins and found that relative solvent accessibility is a stronger predictor of site-wise variability. In addition, the strength of correlations with structural predictors were weaker in \cite{Shahmoradietal2014}.  Here we present research that suggests that differences in the correlations between these two datasets are due to differences in the site variability among proteins within in each dataset. Specifically proteins with a larger variance in entropy among sites exhibit stronger structure-sequence correlations between both local packing density and solvent accessibility. 
\end{abstract}
\vfill
\vfill
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}

@@

Proteins are subject to a number of biophysical and functional constraints \citep{Scherreretal2012, Wilkeetal2010}. These constraints result in site-specific patterns of sequence variability within a protein. Recently several site-specific structural properties that can explain patterns of sequence variability in proteins have been identified. One of the earliest examples was Relative Site Accessibility (RSA). \cite{Fransozaetal2009} identified RSA as the strongest predictor of evolutionary rate and found that residues that are buried within the core of proteins tend to be more conserved than exposed residues close to the surface of the protein. In their analysis, they considered the ability of both RSA and various definitions of residue packing density to predict evolutionary rate. They found that RSA and evolutionary rate shared a significant linear relationship. Afterwards, several other works  also found that RSA as a significant predictor of evolutionary rate and supported this linear relationship \citep{Ramseyetal2011, Scherreretal2012}. However, these papers all have the same flaw. During the course of their analysis they binned the protein sites and averaged over all sites within a bin when determining the trend of RSA. This process may have produced artifacts that account for this strong linear trend between RSA and evolutionary rate. \\
\indent Recently, \cite{Yehetal2014} performed a similar analysis on a series of enzyme monomer proteins and found that packing density, as defined by Contact Number and Weighted Contact Number \citep{Liaoetal2005, Yehetal2014, Huangetal2014}, was the strongest determinant of site variability.  Soon Afterwards, \cite{Shahmoradietal2014} also performed a site-wise analysis on a series of viral proteins. In this analysis they found that RSA had a slightly stronger correlation with site variability as opposed to local packing density.  Moreover, the effect seen between CN and WCN was of a much smaller magnitude as compared to \cite{Yehetal2014}. It is not well understood the reasoning behind the difference in magnitude in the correlations seen between the two studies. Here we attempt to reconcile the work done in this area. We find that site variability is the primary determinant of the strength of structure-sequence relationships and some differences in previous work can be explained in terms of differing levels of site variability. \\

@@

    \subsection*{Structures, sequences, and measures of sequence properties } 
    The results presented in this work are based on two datasets. The first is a dataset of $209$ monomeric enzymes taken from \cite{Huangetal2014}, originally from \cite{Yehetal2014}.The original dataset was comprised of $213$ proteins but we removed four of the proteins (PDB IDS: 1BBS, 1BS0, 1DIN, 2HPL) that had did not have data at insertion sites. Briefly, these proteins are all enzyme monomers   randomly picked from the Catalytic Site Atlas $2.2.11$ \citep{Porteretal2004} with protein sizes in the sample ranging from $95$ to $1287$ residues in length. For each structure we had a corresponding alignment of up to 300 homologous sequences.  The second dataset was taken from \cite{Shahmoradietal2014} and is comprised of nine viral proteins. The viral proteins range from 122 - 557 residues in length and each structure is accompanied by a sequence alignment of up to 2362 homologous sequences. Sequence alignments for both datasets were constructed by aligning the amino-acid sequences using the alignment software MAFFT \citep{Katohetal2002, Katohetal2005}, specifying the auto flag to select the optimal algorithm for the given dataset. The alignments were then used to calculate site-specific measures of sequence variability for each individual protein in both datasets. To do so, we relied on two independent methods of measuring sequence variability. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy at each alignment column $i$:
    \begin{equation}
        \label{eqn:shannon}
        H_i = -\sum_j P_{ij}\ln P_{ij}
    \end{equation}

@@

    \subsection*{Calculation of Structural Properties}
In our analysis we used two types of measures of local packing density used in previous studies: Contact Number (CN) and Weighted Contact Number (WCN). For the purposes of our comparison between the two datasets of interest we used the CN and WCN for the enzyme proteins calculated in \cite{Huangetal2014} and the CN and WCN values for the viral proteins from \cite{Shahmoradietal2014}. In both of these works and in \cite{Yehetal2014}, WCN and CN are both defined the same.  Contact Number is defined as the number of $C_{\alpha}$ within a pre-redefined radius, $r_0$. In this case, $r_0$ = 13 as in the previous papers. Weighed Contact Number for a residue, $i$, is defined as in \cite{Liaoetal2005} and \cite{Huangetal2014} as:

@@

	where $r_j$ is the length between the $C_{\alpha}$ of residue $i$ and residue $j$ in a protein of length $N$ \citep{Yehetal2014}.

@@

We calculate the strength between the structural properties and site variability by calculating the Spearman correlation, $\rho$,  between the structural property (solvent exposure and local packing density) and site variability, as measured by either evolutionary rate or site entropy. On average the correlations between CN and entropy are larger in absolute magnitude in the enzyme proteins as compared to viral proteins. The average $\rho$ between each structural property and each measure of site variability (entropy and evolutionary rates) can be seen in Table \ref{table:entropy_stats}  and Table \ref{table:rate_stats}. The viral proteins experience much lower site variability. Figure 1 shows the correlations between CN and entropy for the viral and enzyme proteins. Even though on average the correlations in the enzyme proteins are larger as reported by \cite{Yehetal2014}, the viral proteins still have correlation strengths that are comparable to some of the enzyme proteins with lower correlations.

@@

\indent In order to further examine the relationship between entropy and the structure-sequence relationships, we used the mean entropy and variance of entropy at sites as predictors of the strength of structure-sequence relationships. Table \ref{table:model_stats} illustrates the coefficients of various linear models.  For WCN-Entropy correlations, mean entropy is not a significant predictor. Although for CN-Entropy correlations, mean entropy is significant the coefficient is 0.079 and therefore predictive power is low. Mean Entropy is also not a significant predictor for the strength of the relationship between RSA and Entropy. For all linear models where mean entropy and dataset were used to predict structure-sequence correlations, dataset was a significant predictor (Table \ref{table:model_stats}). This means that enzyme and virus proteins have different correlations when using mean entropy as a predictor. This agrees with differences seen in the previous works by \cite{Yehetal2014} and \cite{Shahmoradietal2014}.

@@

\section{Discussion}
\label{sec:dcr}
Structure imposes constraints on the evolution of proteins. Therefore structural predictors can be used to predict the variability at sites. Two main determinants of variability at protein sites are solvent accessibility and local packing density. Sites that are on surface of the protein tend to have higher solvent accessibility and exhibit more site variability. Sites that are densely packed and have more contacts tend to evolve slower and exhibit less sequence variability.

\indent Here we examined the relationship between site variability and the strength of structure-sequence relationships. We found that proteins with a larger variance in site variability as measured by sequence entropy, have stronger structure-sequence relationships on average. However, when looking at the mean entropy values of proteins there is a distinct difference between the virus and enzyme datasets. Overall the enzyme proteins have more site variability than the viral proteins.When predicting structure-sequence relationships the variance in site variability among proteins is a better predictor than mean site variability.  Additionally the reason that the enzyme proteins in \cite{Yehetal2014} had higher correlation coefficients is that on average, the enzyme proteins have larger variance in entropy as compared to the virus proteins of \cite{Shahmoradietal2014}. Moreover, since the variance in entropy is a more significant predictor of correlations in local packing density, an increased level of variance in site variability will have a larger effect on the local packing density correlations. It appears that site entropy and evolutionary rate are not equal measures of site variability in terms of predicting the strength of structure-sequence relationships. Finally, the variance in evolutionary rate among sites of a protein has a smaller an effect on the correlation magnitude as compared to entropy. This suggests that the absolute variability at sites is a better predictor of whether structural determinants can be used to predict variability at sites  as opposed to evolutionary rate. Therefore when examining structural determinants of site variability, entropy is a may be a more meaningful quantity.

Maybe change the focus: forget trying to figure out which is the best predictor: RSA or local packing density. Focus on bridging the gap on why there are differences in correlation strength. They had stronger correlations in the Yeh et al (2014) paper as compared to the Shahmoradi paper. Make a table of the mean correlations and variance from both papers.

Proteins are under a number of constraints for function (Liberles et al Paper). Biophysical considerations that have been shown to affect function include: selection for protein stability, specificity for protein binding and selection against extraneous binding. Recently there have been several works that have looked into the effect of protein structure on protein function as well (Xia, Yeh, Shahmoradi, Echave, Huang)

{\color{red} Amir's comment: I think it maybe better to not put so much emphasis on variance of entropy, but its implications and meaning. For example entropy variance is an indicator of sequence divergence, to some extent.}

@@

    \begin{figure}[H]
            \centerline{\includegraphics[width=7.5in]{entropy_cn_cor.pdf}}     
            \caption{Correlations between Contact Number (CN) and Entropy. For each protein we calculate the Spearman Correlation coefficient between CN and entropy at each site within the protein. A) Comparison between the correlation coefficients and mean entropy of site in each protein. The line indicates a linear regression through the enzyme proteins with mean entropy as a single predictor of the spearman correlations. B) Comparison between correlation coefficients and the variance of entropy of each protein. The enzyme proteins are colored in black and the virus proteins are visualized in red. For both WCN and CN the viral proteins have lower mean entropy and variance of entropy. The line indicates a linear regression through the enzyme proteins with variance entropy as a single predictor of the spearman correlations. Proteins with a larger variance in site entropy among sites tend to smaller CN-Entropy correlations in magnitude. }
            \label{fig:entropy_cn_cor}
    \end{figure}

@@

>>>>

\usepackage[margin = 1.0 in]{geometry}
%\bibliographystyle{ynt}
\usepackage{natbib}
%\usepackage[natbib=true,style=numeric,sorting=ynt]{biblatex}

@@

\title{Variation in site entropy explains differences in structure-sequence relationships of proteins}
\author{Eleisha L. Jackson$^{1^*}$, Amir Shahmoradi$^{1*}$, Claus O. Wilke$^{1^{**}}$}
\begin{document}

@@

\noindent
%$^1$ Department of Physics, The University of Texas at Austin, Austin, TX 78712, USA \\
$^1$ Institute of Cellular and Molecular Biology, Center for Computational Biology and Bioinformatics, and Department of Integrative Biology, The University of Texas at Austin, Austin, Texas, 78712 USA\\

@@

\begin{abstract}
Recent work has shown that structural properties are capable of predicting site-specific sequence variability for a given protein. However, the strength and significance of these structure-sequence relations appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. There have two works that have investigated structural predictors of site variability which both present different results based on the strength of correlations of structural predictors. According to \cite{Yehetal2014a}, both solvent accessibility and local packing density are predictors of site-wise variability with local packing density being strongest predictor of site-wise variability. However, recently, \cite{Shahmoradietal2014} compared local packing density, residue flexibility, and solvent accessibility in viral proteins and found that relative solvent accessibility is a stronger predictor of site-wise variability. In addition, the strength of correlations with structural predictors were weaker in \cite{Shahmoradietal2014}.  Here we present research that suggests that differences in the correlations between these two datasets are due to differences in the site variability among proteins within in each dataset. Specifically proteins with a larger variance in entropy among sites exhibit stronger structure-sequence correlations between both local packing density and solvent accessibility. 
\end{abstract}
\vfill
\vfill
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}

@@

The effect of protein-protein interactions(Fraser 2002)

Proteins are subject to a number of biophysical and functional constraints \citep{Scherreretal2012, Wilkeetal2010, Sikoseketal2014, Liberlesetal2009, Huangetal2014, Fraseretal2002, Liaoetal2010, Serohijosetal2012}. These constraints result in site-specific patterns of sequence variability within a protein. Recently several site-specific structural properties that can explain patterns of sequence variability in proteins have been identified. One of the earliest examples was Relative Site Accessibility (RSA). \cite{Fransozaetal2009} identified RSA as the strongest predictor of evolutionary rate and found that residues that are buried within the core of proteins tend to be more conserved than exposed residues close to the surface of the protein. In their analysis, they considered the ability of both RSA and various definitions of residue packing density to predict evolutionary rate. They found that RSA and evolutionary rate shared a significant linear relationship. Afterwards, several other works  also found that RSA as a significant predictor of evolutionary rate and supported this linear relationship \citep{Ramseyetal2011, Scherreretal2012}. However, these papers all have the same flaw. During the course of their analysis they binned the protein sites and averaged over all sites within a bin when determining the trend of RSA. This process may have produced artifacts that account for this strong linear trend between RSA and evolutionary rate. \\
\indent Recently, \cite{Yehetal2014a} performed a similar analysis on a series of enzyme monomer proteins and found that packing density, as defined by Contact Number and Weighted Contact Number \citep{Liaoetal2005, Yehetal2014a, Huangetal2014}, was the strongest determinant of site variability.  Soon Afterwards, \cite{Shahmoradietal2014} also performed a site-wise analysis on a series of viral proteins. In this analysis they found that RSA had a slightly stronger correlation with site variability as opposed to local packing density.  Moreover, the effect seen between CN and WCN was of a much smaller magnitude as compared to \cite{Yehetal2014a}. It is not well understood the reasoning behind the difference in magnitude in the correlations seen between the two studies. Here we attempt to reconcile the work done in this area. We find that site variability is the primary determinant of the strength of structure-sequence relationships and some differences in previous work can be explained in terms of differing levels of site variability. \\

@@

    \subsection*{Structures, sequences, and measures of sequence properties } 
    The results presented in this work are based on two datasets. The first is a dataset of $209$ monomeric enzymes taken from \cite{Huangetal2014}, originally from \cite{Yehetal2014a}.The original dataset was comprised of $213$ proteins but we removed four of the proteins (PDB IDS: 1BBS, 1BS0, 1DIN, 2HPL) that had did not have data at insertion sites. Briefly, these proteins are all enzyme monomers   randomly picked from the Catalytic Site Atlas $2.2.11$ \citep{Porteretal2004} with protein sizes in the sample ranging from $95$ to $1287$ residues in length. For each structure we had a corresponding alignment of up to 300 homologous sequences.  The second dataset was taken from \cite{Shahmoradietal2014} and is comprised of nine viral proteins. The viral proteins range from 122 - 557 residues in length and each structure is accompanied by a sequence alignment of up to 2362 homologous sequences. Sequence alignments for both datasets were constructed by aligning the amino-acid sequences using the alignment software MAFFT \citep{Katohetal2002, Katohetal2005}, specifying the auto flag to select the optimal algorithm for the given dataset. The alignments were then used to calculate site-specific measures of sequence variability for each individual protein in both datasets. To do so, we relied on two independent methods of measuring sequence variability. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy at each alignment column $i$:
    \begin{equation}
        \label{eqn:shannon}
        H_i = -\sum_j P_{ij}\ln P_{ij}
    \end{equation}

@@

    \subsection*{Calculation of Structural Properties}
In our analysis we used two types of measures of local packing density used in previous studies: Contact Number (CN) and Weighted Contact Number (WCN). For the purposes of our comparison between the two datasets of interest we used the CN and WCN for the enzyme proteins calculated in \cite{Huangetal2014} and the CN and WCN values for the viral proteins from \cite{Shahmoradietal2014}. In both of these works and in \cite{Yehetal2014a}, WCN and CN are both defined the same.  Contact Number is defined as the number of $C_{\alpha}$ within a pre-redefined radius, $r_0$. In this case, $r_0$ = 13 as in the previous papers. Weighed Contact Number for a residue, $i$, is defined as in \cite{Liaoetal2005} and \cite{Huangetal2014} as:

@@

	where $r_j$ is the length between the $C_{\alpha}$ of residue $i$ and residue $j$ in a protein of length $N$ \citep{Yehetal2014a}.

@@

Although it is clear that structure plays a role in the evolution of proteins, it is not clear to what extent protein structure constrains sequence variability. To date there have been conflicting reports about the strength of structure-sequence correlations \citep{Yehetal2014b, Shahmoradietal2014}. Therefore we sought to determine the main determinants of structure-sequence correlations within proteins. First, we calculated the strength between the structural properties and site variability by calculating the Spearman correlation, $\rho$,  between the structural property (solvent exposure and local packing density) and site variability, as measured by either evolutionary rate or site entropy. On average the correlations between CN and entropy are larger in absolute magnitude in the enzyme proteins as compared to viral proteins. The average $\rho$ between each structural property and each measure of site variability (entropy and evolutionary rates) can be seen in Table \ref{table:entropy_stats}  and Table \ref{table:rate_stats}. The viral proteins experience much lower site variability. Figure 1 shows the correlations between CN and entropy for the viral and enzyme proteins. Even though on average the correlations in the enzyme proteins are larger as reported by \cite{Yehetal2014a}, the viral proteins still have correlation strengths that are comparable to some of the enzyme proteins with lower correlations.

@@

\indent In order to further examine the relationship between entropy and the structure-sequence relationships, we used the mean entropy and variance of entropy at sites as predictors of the strength of structure-sequence relationships. Table \ref{table:model_stats} illustrates the coefficients of various linear models. For WCN-Entropy correlations, mean entropy is not a significant predictor. Although for CN-Entropy correlations, mean entropy is significant the coefficient is 0.079 and therefore predictive power is low. Mean Entropy is also not a significant predictor for the strength of the relationship between RSA and Entropy. For all linear models where mean entropy and dataset were used to predict structure-sequence correlations, dataset was a significant predictor (Table \ref{table:model_stats}). This means that enzyme and virus proteins have different correlations when using mean entropy as a predictor. This agrees with differences seen in the previous works by \cite{Yehetal2014a} and \cite{Shahmoradietal2014}.

@@

\section{Discussion}
\label{sec:dcr}

There has been previous work that has illuminated several factors that determine the rate at which proteins evolve \citep{Fraseretal2002, Drummondetal2008, Wilkeetal2010, Yangetal2012}. Some of these factors include: expression level \citep{Drummondetal2008, Wilkeetal2010, Paletal2001, Subramanianetal2004}, interactions with other protein partners \citep{Fraseretal2002, Yangetal2012, Minterisetal2005, Pangetal2010}  and selection for the costs of misfolding \citep{Drummondetal2005}.

\indent Proteins are under a number of constraints for function \citep{Liberlesetal2009}. Biophysical considerations that have been shown to affect function include: selection for protein stability and proper folding \citep{Drummondetal2005}, specificity for protein binding \citep{Zarrinparetal2003}  and selection against non-specific and extraneous binding \citep{Levyetal2012}, selection to minimize protein aggregation \citep{DePristoetal2005}  and protein structure \citep{Fransozaetal2009, Fransozaetal2012, Yehetal2014a, Yehetal2014b, Shahmoradietal2014, Echaveetal2015, Huangetal2014}. Recently there have been several works that have looked into the effect of protein structure on protein function. Local packing density and solvent accessibility have been emerged as two structural predictors that have been showed to predict site variability \citep{Yehetal2014a, Yehetal2014b, Shahmoradietal2014, Fransozaetal2009, Bustamanteetal2000}.

\indent Structure imposes constraints on the evolution of proteins. Therefore structural predictors can be used to predict the variability at sites. Two main determinants of variability at protein sites are solvent accessibility and local packing density. Sites that are on surface of the protein tend to have higher solvent accessibility and exhibit more site variability. Sites that are densely packed and have more contacts tend to evolve slower and exhibit less sequence variability. However, the strength of how well these two structural quantities (local packing density and solvent accessibility) has not been clearly explained. \cite{Yehetal2014a} examined the effect of using solvent accessibility and local packing density as measures of predicting site-wise variability. They found that both were predictors. However, \cite{Shahmoradietal2014} performed the same analysis on viral proteins and found that will both were predictors of stir-wise variability, the correlations between these two structural predictors and site variability were significantly lower.

\indent Here we examined the relationship between site variability and the strength of structure-sequence relationships in order to dissect the nature of the correlation between structural properties and site variability. We found that proteins with a larger variance in site variability as measured by sequence entropy, have stronger structure-sequence relationships on average. However, when looking at the mean entropy values of proteins there is a distinct difference between the virus and enzyme datasets. Overall the enzyme proteins have more site variability than the viral proteins. When predicting structure-sequence relationships the variance in site variability among proteins is a better predictor than mean site variability.  \indent Additionally the reason that the enzyme proteins in \cite{Yehetal2014a} had higher correlation coefficients is that on average, the enzyme proteins have larger variance in entropy as compared to the virus proteins of \cite{Shahmoradietal2014}. Moreover, since the variance in entropy is a more significant predictor of correlations in local packing density, an increased level of variance in site variability will have a larger effect on the local packing density correlations. It appears that site entropy and evolutionary rate are not equal measures of site variability in terms of predicting the strength of structure-sequence relationships. Finally, the variance in evolutionary rate among sites of a protein has a smaller an effect on the correlation magnitude as compared to entropy. This suggests that the absolute variability at sites is a better predictor of whether structural determinants can be used to predict variability at sites  as opposed to evolutionary rate. Therefore when examining structural determinants of site variability, entropy is a may be a more meaningful quantity.

@@

    \begin{figure}[H]
            \centerline{\includegraphics[width=7.5in]{entropy_cn_cor.pdf}}     
            \caption{Correlations between Contact Number (CN) and Entropy. For each protein we calculate the Spearman Correlation coefficient between CN and entropy at each site within the protein. A) Comparison between the correlation coefficients and mean entropy of site in each protein. The line indicates a linear regression through the enzyme proteins with mean entropy as a single predictor of the spearman correlations. B) Comparison between correlation coefficients and the variance of entropy of each protein. The enzyme proteins are colored in black and the virus proteins are visualized in red. For both WCN and CN the viral proteins have lower mean entropy and variance of entropy. The line indicates a linear regression through the enzyme proteins with variance entropy as a single predictor of the spearman correlations. Proteins with a larger variance in site entropy among sites tend to smaller CN-Entropy correlations in magnitude. Note: The entropy values in \cite{Shahmoradietal2014} were scaled.}
            \label{fig:entropy_cn_cor}
    \end{figure}

@@

Make a table of the mean correlations and variance from both papers.
